---
title: "Success of Medical Management of Tubal Ectopic Pregnancy with Methotrexate"
author: "Osama Ahmed Ibrahim"
output: word_document
---

```{r setup, 	include = FALSE}
knitr::opts_chunk$set(
	message = FALSE,
	warning = FALSE,
	dpi = 300
)
```
### Statistical Analysis
Descriptive statistics were reported as median (interquartile range, Q1–Q3) for continuous variables and as frequency (percentage) for categorical variables. Univariable logistic regression was performed to evaluate the association of each candidate predictor with methotrexate treatment outcome, reported as unadjusted odds ratios (OR) with 95% confidence intervals (CI). Two multivariable logistic regression models were subsequently constructed: Model 1 included BMI, gestational age (weeks), b-HCG, and presence of free fluid; Model 2 included BMI, b-HCG, and free fluid. Nested model comparison was performed using the likelihood ratio test via analysis of deviance (χ² test). Model calibration was assessed using the Hosmer–Lemeshow goodness-of-fit test with 10 groups, where a non-significant p-value (> 0.05) indicated adequate calibration. Model fit was further quantified using McFadden's, Nagelkerke's, and Cox–Snell pseudo-R² statistics. Discrimination was assessed by the area under the receiver operating characteristic curve (AUC) with 95% CI, and the two models were formally compared using DeLong's test for correlated ROC curves. Multicollinearity was assessed using the variance inflation factor (VIF), with values < 5 considered acceptable. A clinical nomogram was constructed from the final model using the rms package to enable individual patient probability estimation. All analyses were performed in R (version 4.x) using the gtsummary, pROC, ResourceSelection, DescTools, and rms packages. A two-tailed p-value < 0.05 was considered statistically significant.

```{r message=FALSE, warning=FALSE, include=FALSE}
library(readxl)
library(tidyverse)
library(gtsummary)
library(patchwork)
library(rstatix)
library(car)
library(corrplot)
library(DescTools)
library(ResourceSelection)
library(pROC)
library(rms)
library(flextable)


ectopic_pregnancy_data <- read_excel(path = "Project 5 data.xlsx")

ectopic_pregnancy_data <- ectopic_pregnancy_data %>% rename_with(str_to_title, 
                                                                 c(-BMI, -ID)) %>% 
  rename_with(str_to_upper, Bhcg) %>% 
  rename("EP Mass Diameter(cm)" = `Ep Mass Diameter(Cm)`,
         "Number of Doses" = `No. Of Doses`)


ectopic_pregnancy_data <- ectopic_pregnancy_data %>% mutate(across(
  c(`Free Fluid`, Outcome, `Number of Doses`), ~ factor(str_to_title(.x))
))
  

ectopic_pregnancy_data <- ectopic_pregnancy_data %>% mutate(
  `Number of Doses` = factor(`Number of Doses`, levels = 
                               c("One", "Two", "Multiple"))
)
```

### Results
### Study Population
A total of 121 women with tubal ectopic pregnancy who received methotrexate were included. The median age was 35 years (IQR: 30–39) and the median BMI was 28.5 kg/m² (IQR: 26.5–30.8). The median gestational age at treatment was 6.20 weeks (IQR: 5.50–7.00), and the median serum b-HCG level was 2,569 mIU/mL (IQR: 1,187–4,560). Free fluid was detected in 17 patients (14%). The median ectopic mass diameter was 2.60 cm (IQR: 1.70–3.50). Baseline characteristics are summarized in Table 1.


#### Table 1. Baseline characteristics of patients with tubal ectopic pregnancy managed with methotrexate
```{r echo=FALSE, message=FALSE, warning=FALSE}
ectopic_pregnancy_data %>% tbl_summary(include = -ID,
                                       type = list(
                                         Gravity ~ "continuous",
                                         Parity ~ "continuous",
                                         Abortion ~ "continuous"
                                       ),
                                       label = list(
                                         BHCG ~ "b-HCG (mIU/mL)",
                                         `EP Mass Diameter(cm)` ~ "Ectopic mass diameter (cm)"
                                       )) %>% modify_footnote_header("Abbreviations: BMI = body mass index; b-HCG = beta human chorionic gonadotropin.",
                       columns = "label") %>% 
  bold_labels() %>% as_flex_table()
```


### Treatment Outcome

Methotrexate treatment was successful in 99 patients (82%), while treatment failed in 22 patients (18%). Among those who succeeded, 81 patients (67%) required a single dose, 29 (24%) required two doses, and 11 (9.1%) required multiple doses.

Univariable logistic regression identified several candidate predictors of methotrexate failure. On multivariable analysis, both models demonstrated adequate fit. The likelihood ratio test comparing Model 1 (BMI, gestational age, b-HCG, free fluid) to Model 2 (BMI, b-HCG, free fluid) showed no statistically significant difference in model deviance (χ² = −0.233, df = 1, p = 0.630), indicating that gestational age did not contribute meaningfully beyond the other predictors. The adjusted odds ratios for both models are presented in Table 2.

#### Table 2. Univariable and multivariable logistic regression analysis of predictors of methotrexate treatment failure.
```{r echo=FALSE, message=FALSE, warning=FALSE}
univariable_reg_table <- ectopic_pregnancy_data %>% select(Age, BMI, Gravity, Parity, Abortion, `Gestational Age (Weeks)`,
                                  BHCG, `Free Fluid`, `EP Mass Diameter(cm)`, Outcome) %>% 
  tbl_uvregression(label = list(
    BHCG ~ "b-HCG (mIU/mL)",
    `EP Mass Diameter(cm)` ~ "Ectopic mass diameter (cm)"
  ), method =  "glm", y = Outcome,
  method.args = list(family = "binomial"), exponentiate = T,
  hide_n = T) %>% bold_labels()


multivariable_model1 <- glm(Outcome ~ BMI + `Gestational Age (Weeks)` + BHCG + `Free Fluid`, 
                               family = "binomial", data = ectopic_pregnancy_data)

multivariable_model2 <- glm(Outcome ~ BMI + BHCG + `Free Fluid`, family = "binomial", 
                            data = ectopic_pregnancy_data)

table_multivaraible_1 <- tbl_regression(multivariable_model1, exponentiate = T) %>% bold_p(t = 0.05) %>% 
  bold_labels()

table_multivaraible_2 <- tbl_regression(multivariable_model2, exponentiate = T) %>% bold_p(t = 0.05) %>% 
  bold_labels()
tbl_merge(tbls = list(univariable_reg_table, table_multivaraible_1,
                      table_multivaraible_2),
          tab_spanner = c("**Unadjusted**", "**Model 1**", "**Model 2**")) %>% 
  # Footnote on the OR column headers
  modify_footnote_header(
    footnote = "OR = odds ratio; CI = confidence interval. Exponentiated logistic regression coefficients.",
    columns  = starts_with("estimate")
  ) %>%
  # Footnote on the spanning header for Model 1
  modify_footnote_spanning_header(
    footnote = "Model 1: BMI + gestational age + b-HCG + free fluid",
    columns  = "estimate_2"   # first column under Model 1 spanner
  ) %>%
  # Footnote on the spanning header for Model 2
  modify_footnote_spanning_header(
    footnote = "Model 2: BMI + b-HCG + free fluid",
    columns  = "estimate_3"   # first column under Model 2 spanner
  ) %>%
  # Footnote on a specific body row — e.g., flag free fluid as binary
  modify_footnote_body(
    footnote = "Binary variable: Yes = free fluid present on ultrasound",
    columns  = "label",
    rows     = variable == "Free Fluid" & row_type == "label"
  ) %>%
  as_flex_table() %>% set_table_properties(layout = "autofit", width = 1) %>% autofit()
```

### Model Performance
The Hosmer–Lemeshow test for Model 2 yielded a non-significant result (χ² = 11.40, df = 8, p = 0.180), indicating good calibration between observed and predicted outcomes. Pseudo-R² statistics for Model 2 were: McFadden = 0.407, Nagelkerke = 0.523, and Cox–Snell = 0.320, collectively indicating excellent model fit. Both Model 1 and Model 2 achieved identical discrimination, with an AUC of 0.921 (95% CI reported in Figure 2). DeLong's test confirmed no significant difference in discriminative performance between the two models (Z = 0, p = 1.000). All VIF values for Model 1 were below 1.20 (BMI = 1.19, gestational age = 1.05, b-HCG = 1.09, free fluid = 1.17), confirming the absence of multicollinearity.

#### Figure 1. Receiver operating characteristic (ROC) curves for Model 1 and Model 2 predicting methotrexate treatment success.

```{r echo=FALSE, fig.height=5, fig.width=7, message=FALSE, warning=FALSE}

roc1 <- roc(multivariable_model1$y, fitted(multivariable_model1), ci = TRUE)
roc2 <- roc(multivariable_model2$y, fitted(multivariable_model2), ci = TRUE)
# Overlay plot
plot(roc1, col = "steelblue", lwd = 2, legacy.axes = TRUE,
     main = "ROC Curve Comparison")
# add argument make the curve added to an existing plot
plot(roc2, col = "tomato", lwd = 2, add = TRUE)

legend("bottomright",
       legend = c(
         paste0("Model 1 (AUC = ", round(auc(roc1), 3),
                ", 95% CI: ", round(ci.auc(roc1)[1], 3), "–", round(ci.auc(roc1)[3], 3), ")"),
         paste0("Model 2 (AUC = ", round(auc(roc2), 3),
                ", 95% CI: ", round(ci.auc(roc2)[1], 3), "–", round(ci.auc(roc2)[3], 3), ")")
       ),
       col = c("steelblue", "tomato"), lwd = 2)
```


#### Figure 2. Nomogram for individualized prediction of methotrexate treatment success in tubal ectopic pregnancy.
```{r echo=FALSE, fig.height=7, fig.width=10, message=FALSE, warning=FALSE}
cleaned_name_data <- ectopic_pregnancy_data %>% rename(
  "Free_Fluid" = `Free Fluid`
)
  
# Step 1: Tell rms about your data distributions
dd <- datadist(cleaned_name_data)
options(datadist = "dd")

# Step 2: Refit your model using lrm() instead of glm()
lrm_model <- lrm(Outcome ~ BMI + BHCG + Free_Fluid,
                 data = cleaned_name_data, x = TRUE, y = TRUE)

# Step 3: Build the nomogram
nom <- nomogram(lrm_model,
                fun = plogis, # converts log-odds to probability
                fun.at = c(0.1, 0.3,0.5,0.7,0.9),
                funlabel = "Predicted Probability")

# Step 4: Plot it
plot(nom, xfrac    = 0.4,
     cex.axis = 1.2,   # tick mark label size
     cex.var  = 1.3,   # variable name label size (left side)
     lmgp     = 0.4)
```

